Competing Constructions for German Measure Noun Phrases: from Usage Data to Experimental Validation

### 2017-04-01 Latex/article.tex
milestone: done except intro/conclusion

\usepackage[small]{dgruyter}
\usepackage{microtype}
\usepackage{enumerate}
\usepackage{eqnarray}

\newcommand{\wrt}{w.\,r.\,t.\ }
\newcommand{\Wrt}{W.\,r.\,t.\ }
\newcommand{\eg}{e.\,g.,}
\newcommand{\Eg}{E.\,g.,}
\newcommand{\ie}{i.\,e.,}
\newcommand{\Ie}{I.\,e.,}
\newcommand{\Sub}[1]{\ensuremath{\mathrm{_{#1}}}}
\newcommand{\Sup}[1]{\ensuremath{\mathrm{^{#1}}}}
\newcommand{\Subsf}[1]{\ensuremath{\mathsf{_{#1}}}}
\newcommand{\Supsf}[1]{\ensuremath{\mathsf{^{#1}}}}
\newcommand{\pPB}{p\Subsf{PB}}

@@

@@

\usepackage{bbding}
\newcommand{\CheckIt}{\CheckmarkBold}

@@

  \author*[1]{Roland Schäfer}
  %\runningauthor{...}
  \affil[1]{Freie Universität Berlin}
  \title{Competing Constructions for German Measure Noun Phrases: from Usage Data to Experimental Validation}
  \runningtitle{Competing Constructions for German Measure Noun Phrases}
  %\subtitle{...}
  \abstract{}
  \keywords{corpus methods and experimental methods, alternations, hierarchical models, measure constructions, German}
  %\classification[PACS]{...}
  %\communicated{...}
  %\dedication{...}
  %\received{...}
  %\accepted{...}
  \journalname{(submitted)}
  \journalyear{2017}
  %\journalvolume{...}
  %\journalissue{...}
  %\startpage{...}
  %\aop
  %\DOI{...}

@@

\section{Cognitively oriented corpus linguistics: theories, case studies, and methods}
\label{sec:cogocl}

@@

The alternation phenomenon to be discussed in this paper appears to be a straightforward case alternation at first sight.
However, without a solid knowledge of German nominal morpho-syntax, some of the more subtle distinctions that have to be made in its analysis can be confusing.
Therefore, I introduce and illustrate the relevant constructions in some detail in this section.
I describe the narrow range of syntactic constellations in which the alternation occurs, and I motivate the focus on \textit{only} this narrow (range rather than, for example, the whole range of nominal constructions expressing quantities).

I use the term \textit{measure noun phrase} (MNP) to refer a noun phrase (NP) in which a kind-denoting (count or mass) noun depends on another noun specifying a quantity of the objects or the substance denoted by the kind noun.
I call the kind-denoting noun simply the \textit{kind noun} and the quantity-denoting noun the \textit{measure noun}.
Measure nouns can be all sorts of nouns which genuinely denote a quantity (such as \textit{litre} or \textit{amount}) but also nouns denoting containers, collections, etc. (such as \textit{glass} or \textit{bucket}).
In a similar vein, \citet[284]{Brems2003} considers nouns as measure nouns \textit{which, strictly speaking, do not designate a `measure', but display a more nebulous} (\textit{sic!}) \textit{potential for quantification} (see also \citealp[530]{Koptjevskaja2001}, and \citealp[338]{Rutkowski2007}).
For illustration purposes, in the English \textit{a cup of fine coffee}, \textit{cup} is the measure noun, and \textit{coffee} is the kind noun.

In the case at hand, three different syntactic configurations need to be distinguished \wrt case assignment inside German measure noun phrases.
If the kind noun forms an NP with a determiner, the construction resembles (and is usually called) a \textit{pseudo-partitive} (on partitives and pseudo-partitives see, e.\,g., \citealp{Barker1998,Selkirk1977,Stickney2007,Vos1999}; for a recent application of the terminology to German, see \citealp{Gerstenberger2015}).%
\footnote{If the kind noun is definite, the construction instantiates a true partitive.
Whereas partitives are constructions denoting a proper part-of relation strictly requiring a definite kind noun as in \textit{a sip of the wine}, pseudo partitives -- albeit syntactically similar and diachronically related to partitives in many languages -- merely denote quantities and contain indefinite kind nouns as in \textit{a sip of wine}.
%In the literature on German, some authors incorrectly call the pseudo-partitive a \textit{partitive} \cite{Hentschel1993} while some realise the difference and at least mention it \citep{Eschenbach1994,GallmannLindauer1994,Loebel1989,Zimmer2015}.
}
Here, the kind noun is in the genitive, and I refer to the construction in (\ref{ex:intro:pseudopartitive1}) as the \textit{Pseudo-partitive Genitive Construction} (PGC).

@@

When the kind noun is bare -- \ie, without a determiner and without modifying adjectives -- the kind noun has to agree in case with the measure noun, and the genitive seen in the PGC is not acceptable, see (\ref{ex:intro:narrowapposition}).
This construction is usually classified as a \textit{Narrow Apposition Construction} \citep{Loebel1986}, henceforth NAC.%
\footnote{Notice that the unavailability of the genitive on the kind noun can be seen as following from a rather quirky constraint that genitive NPs in German require the presence of some strongly case-marked element (determiner or adjective) in addition to the head noun in order to be acceptable \citep{GallmannLindauer1994,Schachtl1989}.}
The construction as in (\ref{ex:intro:narrowapposition2}) is also referred to as the \textit{Direct Partitive Construction} for other Germanic languages in which the PGC with the synthetic genitive is not available.%
\footnote{This nomenclature makes sense in contrast to the \textit{Indirect Partitive Construction} with prepositional linkers translating to `of' -- \ie\ analytic genitives -- in such languages, see \cite{HankamerMikkelsen2008} for Danish.
For German, this terminology is not distinctive enough, which is why I use the terms NAC and PGC.}
While only an accusative is shown in (\ref{ex:intro:narrowapposition}), the kind noun obligatory agrees in case also with nominative and dative measure nouns.

@@

The actual alternation can be observed \textit{only when the kind noun occurs with an attributive adjective but without a determiner}, as in (\ref{ex:intro:alternation}), where both the NAC in (\ref{ex:intro:alternation1}) and the PGC in (\ref{ex:intro:alternation2}) are equally acceptable.
They are by-and-large functionally and semantically equivalent.
However, Section~\ref{sec:analyses} is devoted to developing hypotheses about subtle differences between them.%
\footnote{Some descriptive and normative grammars take stronger positions w.\,r.\,t.\ the acceptability of the two options.
See \cite{Hentschel1993,Zimmer2015} for analyses of the sometimes absurd stances taken in grammars of German.
%As the usage and experimental data presented below should render it clear, there might be preferences under certain circumstances, but we cannot assume either construction to be unacceptable.
}

@@

\begin{table}
  \centering
  \begin{tabular}{lccc}
    \multicolumn{1}{r}{kind NP is:} & bare noun NP & NP with adjective & NP with determiner \\
    & [\ldots N\Subsf{meas} [N\Subsf{kind}]] & [\ldots N\Subsf{meas} [AP N\Subsf{kind}]] & [\ldots N\Subsf{meas} [D N\Subsf{kind}]] \\
    \midrule
    narrow apposition         & \CheckIt & \CheckIt &          \\
    pseudo-partitive genitive &          & \CheckIt & \CheckIt \\
  \end{tabular}
  \caption{Distribution of the NAC and PGC constructions in different NP structures}
  \label{tab:constructions}
\end{table}

The distribution of the case patterns in the NAC (case identity between the measure noun and the kind noun) and in the PGC (genitive on the kind noun) depending on the structure of the kind NP are summarised in Table~\ref{tab:constructions}.
I call the narrow apposition construction with a bare kind noun the \NACb, the partitive genitive with a determiner in the kind noun phrase the \PGCd.
I call the alternating variants with an adjective but no determiner in the kind noun phrase \NACa, and \PGCa, respectively.

@@

\begin{enumerate}[i.]
  \item{\label{it:intro:idiot} alleged alternatives to the NAC and the PGC}
  \item{\label{it:intro:datsg} alternative \textit{weak} forms of dative singular neuter adjectives}
  \item{\label{it:intro:femsg} case syncretism in feminine NPs}
  \item{\label{it:intro:plurl} similar constructions with plural/collective kind nouns}
  \item{\label{it:intro:noifl} grammaticalised non-inflected measure nouns}
  \item{\label{it:intro:other} alternative constructions for expressing quantities}
\end{enumerate}

@@

As for (\ref{it:intro:datsg}), a complication with neuter kind nouns in the dative is mentioned by \citet[20--22]{Zimmer2015}.
He reports a high number of occurrences of adjectives being inflected according to the \textit{weak} inflectional pattern, which is normally only used if a strongly inflected determiner precedes the adjective: \textit{kalt-en Wasser} `cold water' instead of \textit{kalt-em Wasser}.%
\footnote{See Section~\ref{sec:analyses} for more discussion of adjectival inflection.}
In my corpus sample, this tendency was not nearly as clear, and there was a high number of very noisy sentences among those potentially showing this pattern.
Hence, I do not discuss these forms much, although I will offer a possible interpretation in Section~\ref{sec:analyses}.

\label{page:femininesyncretism}
With feminine kind nouns, (\ref{it:intro:femsg}) is relevant.
Whereas the singular masculine and neuter nominal subsystem is a four case system (at least for NPs with adjectives and determiners), feminine singular NPs show nominative-accusative and dative-genitive syncretisms.
Table~\ref{tab:syncretisms} shows this as well as the partial syncretism in the plural, where no gender distinctions are made.
This means that with dative measure nouns, the alternation is unobservable if the kind noun is feminine, and it is problematic to speak of a \textit{genitive} kind noun construction in this case.
This is taken into account in the statistical modelling and the discussion presented in Section~\ref{sec:corpusstudies}.

\begin{table}
  \centering
  \begin{tabular}{llll}
     & Masculine\slash Neuter & Feminine & Plural \\
     \midrule
     nominative & rot-er Wein    & \multirow{2}{*}{frisch-e Sahne}   & \multirow{2}{*}{klein-e Äpfel} \\
     accusative & rot-en Wein    &                                   &                                \\
     dative     & rot-em Wein    & \multirow{2}{*}{frisch-er Sahne}  & klein-en Äpfel-n               \\
     genitive   & rot-en Wein-es &                                   & klein-er Äpfel                 \\
  \end{tabular}
  \caption{Case syncretisms in NPs with an adjective and without a determiner (\textit{roter Wein} `red wine', \textit{frische Sahne} `fresh cream', \textit{kleine Äpfel} `small apples')}
  \label{tab:syncretisms}
\end{table}

Turning to (\ref{it:intro:plurl}), readers might have noticed that so far only MNPs denoting quantities of substances (mass kind nouns such as \textit{ein Glas roter Wein\slash roten Weines} `a glass of red wine') have been discussed.
If the kind noun is a plural count noun as in \textit{ein Sack kleine Äpfel\slash kleiner Äpfel} `a bag of small apples' (see Table~\ref{tab:syncretisms} for the inflection pattern), a similar alternation between PGC and NAC can be observed.
%\footnote{\label{fn:eschenbash}On the collective interpretation of such constructions, see \cite{Eschenbach1994}.
%It is, however, unfortunate that Eschenbach does not notice and discuss the morpho-syntactic distribution of PGC and NAC, even when she gets very close to listing the relevant data \citep[217]{Eschenbach1994}.}
In line with experimental results reported in \citet[15--16]{Zimmer2015}, I found that the PGC is so dominant with plural kind nouns (67 of 861 cases or over 92\%, cf.\ Section~\ref{sec:corpusstudies}) that the alternation cannot be analysed in the same way as in the singular case.
While this will play a role in the interpretation of the corpus findings, MNPs with plural ve kind nouns will not be dealt with prominently in the remainder of the paper.
%A statistical model for plural kind nouns will be reported in passing in Section~\ref{sec:corpusstudies}, however.

As for (\ref{it:intro:noifl}), some measure nouns have been grammaticalised in a way that they always appear non-inflected.
They are typical measure nouns like \textit{Gramm} `gram', \textit{Pfund} `pound' or \textit{Prozent} `percent', which do not have plural forms at all.%
\footnote{With neglectable frequency, this effect is extended to container nouns like \textit{Glas} `glass', but with a twist.
A non-inflected variant \textit{zwei Glas roter Wein} `two glasses of wine' and an inflected variant \textit{zwei Gläser roter Wein} co-exist.
The non-inflected variant denotes the quantity of wine fitting into two glasses, the inflected variant denotes two glasses filled with wine.}
I treat these cases like any other measure noun because they enter into both the \NACa\ as in (\ref{ex:uninflectedmeasures:a}) and the \PGCa\ as in (\ref{ex:uninflectedmeasures:b}).
In Section~\ref{sec:analyses}, degrees of grammaticalisation as a factor influencing the alternation will be discussed, however.

@@

Finally, (\ref{it:intro:other}) suggests that there are alternative ways of expressing similar quantificational meanings.
%First of all, the analytic (pseudo-)partitive with \textit{von} `of' is only available as an alternative to the PGC if the kind noun phrase contains a (definite or indefinite) determiner as in (\ref{ex:analyticalpartitive}).
%
%\begin{exe}
%  \ex\label{ex:analyticalpartitive} 
%  \begin{xlist}
%    \ex[ ]{\label{ex:analyticalpartitive:a}\gll ein Glas von dem roten Wein\\
%                                                a glass of the red wine\\
%                                                \glt a glass of the red wine
%       }
%    \ex[*]{\label{ex:analyticalpartitive:b}\gll ein Glas von rotem Wein\\
%                                                a glass of red wine\\
%                                                \glt a glass of red wine
%       }
%  \end{xlist}
%\end{exe}
%
If the measure noun denotes a container, constructions with \textit{voll} (\textit{von})\slash\textit{voller} `full (of)' and \textit{mit} `with' are available, see (\ref{ex:alternatives}).

@@

These construction have very idiosyncratic properties (the construction with \textit{voller} is discussed in \citealp{Zeldes2018}, the construction with \textit{mit} is discussed in \citealp{Bhatt1990}), they are not semantically equivalent to the NAC and the PGC, and -- most importantly -- they are only available for a small subset of measure nouns.
Subtle interactions between these constructions and NAC\slash PGC cannot be excluded, especially because the former might act as a (partial) alternative and thus reduce the count of exemplars of the NAC, the PGC, or both for container measure nouns.
There are no clear hypotheses about such interactions, however.
Nevertheless, the corpus-based models in Section~\ref{sec:corpusstudies} are specified in a way that they could detect preferences for either NAC or PGC for the class of container measure nouns, for which the alternative constructions illustrated in (\ref{ex:alternatives}) are available.

@@

This section briefly reviews existing analyses of the \PGCa-\NACa\ alternation and related issues.
I also develop my own analysis and the appropriate hypotheses for the empirical studies presented in Sections~\ref{sec:corpusstudies} and \ref{sec:externalvalidation}.
I discuss four main factors that might influence the alternation:
first, the different syntactic structures of the variants \NACa\ and \PGCa\ in relation to their syntactic prototypes \NACb\ and \PGCd;
second, degrees of grammaticalisation of the measure noun associated with the two prototypes;
third, a semantic prototypicality effect of the degrees of grammaticalisation (plurality);
fourth, preferences in different registers.

The first aspect is related to analyses of the morpho-syntax of German MNPs.
A major interest in descriptive and generative linguistics is to find a plausible analysis of the hierarchical syntactic structure of the constructions in question.
The first question is which noun constitutes the head of the whole construction.
This question was answered for the case at hand quite clearly by \cite{Loebel1986} already (see also \citealp[213]{Eschenbach1994}, and \citealp[16]{GallmannLindauer1994}).
Subject-verb agreement is always realised on the measure noun, and we can therefore assume that the measure noun is the head.
However, the MNP-internal structure has more interesting cues to offer if the strong\slash weak inflection patterns of adjectives (already mentioned in Section~\ref{sec:descriptive}) are taken into account.
In NPs with a strongly inflected determiner, attributive adjectives inflect according to the massively syncretistic \textit{weak} pattern.
%\footnote{The weak pattern has only two suffixes.
%The structural cases (nominative and accusative) in the singular are marked with \textit{-e}, the oblique cases (dative and genitive) as well as the entire plural are marked with \textit{-en}.
%The only exception to this rule is the masculine accusative singular, which is also marked with \textit{-en}.}
If there is no determiner (as is the case in the alternating constructions), attributive adjectives inflect like determiners, however.
This is called the \textit{strong} inflectional pattern (see Table~\ref{tab:syncretisms}).
Thus, the adjectives in the \NACa\ and the \PGCa\ have properties of adjectives as well as determiners.
They are lexical adjectives and function as attributive modifiers.
On the other hand, they are inflected like determiners, and they are the leftmost element in the NP, which is typical of determiners.
This unusual double nature of adjectives in NPs without determiners leads to a plausible probabilistic interpretation of the pattern shown in Table~\ref{tab:constructions}.
Whenever speakers classify the adjective in the kind noun phrase more as a determiner, they have to use the \PGCa\ because, if there is a determiner, the PGC (just like in the case of the \PGCd) is the only option.
When they classify the adjective more as an adjective, the kind NP has no determiner, and they have to use the \NACa\ (like in the case of the \NACb).%
\footnote{While the generative analysis presented in \cite{Bhatt1990} cannot properly deal with probabilistic effects, Bhatt comes close to this interpretation by analysing the kind NP in the GPC as a DP and in the NAC as an NP.}
In a framework that assumes similarity-based probabilistic classification, such as Prototype Theory, we can regard the \PGCd\ as instantiating the PGC prototype and the \NACb\ as the NAC prototype.
The choice between \PGCa\ and \NACa\ is then expected to be influenced by the degree of similarity to either of these prototypes.

>>>>

\usepackage[small]{dgruyter}
\usepackage{microtype}
\usepackage{enumerate}
\usepackage{eqnarray}
\usepackage{colortbl}

\newcommand{\wrt}{w.\,r.\,t.\ }
\newcommand{\Wrt}{W.\,r.\,t.\ }
\newcommand{\eg}{e.\,g.,}
\newcommand{\Eg}{E.\,g.,}
\newcommand{\ie}{i.\,e.,}
\newcommand{\Ie}{I.\,e.,}
\newcommand{\Sub}[1]{\ensuremath{\mathrm{_{#1}}}}
\newcommand{\Sup}[1]{\ensuremath{\mathrm{^{#1}}}}
\newcommand{\Subsf}[1]{\ensuremath{\mathsf{_{#1}}}}
\newcommand{\Supsf}[1]{\ensuremath{\mathsf{^{#1}}}}
\newcommand{\pPB}{p\Subsf{PB}}
\newcommand{\mpPB}{\ensuremath{p_{\text{PB}}}}

@@

\definecolor{lg}{rgb}{.8,.8,.8}
\newcommand{\Dim}{\cellcolor{lg}}

@@

\usepackage{bbding}
\newcommand{\CheckIt}{\CheckmarkBold}
\newcommand{\CrossIt}{\XSolidBold}

@@

  \author*[1]{Roland Schäfer}
  %\runningauthor{...}
  \affil[1]{Freie Universität Berlin}
  \title{Competing Constructions for German Measure Noun Phrases: from Usage Data to Experimental Validation}
  \runningtitle{Competing Constructions for German Measure Noun Phrases}
  %\subtitle{...}
  \abstract{}
  \keywords{corpus methods and experimental methods, self-paced reading, forced choice, alternations, hierarchical models, measure constructions, German}
  %\classification[PACS]{...}
  %\communicated{...}
  %\dedication{...}
  %\received{...}
  %\accepted{...}
  \journalname{(submitted)}
  \journalyear{2017}
  %\journalvolume{...}
  %\journalissue{...}
  %\startpage{...}
  %\aop
  %\DOI{...}

@@

\section{Cognitively oriented corpus linguistics}
\label{sec:cogocl}

@@

In this section, I introduce and illustrate the relevant alternating constructions.
I describe the narrowly defined syntactic configuration in which the alternation occurs, and I motivate the focus on \textit{only} this narrow (range rather than, for example, the whole range of nominal constructions expressing quantities).

I use the term \textit{measure noun phrase} (MNP) to refer a noun phrase (NP) in which a kind-denoting (count or mass) noun depends on another noun that specifies a quantity of the objects or the substance denoted by the kind noun.
I call the kind-denoting noun simply the \textit{kind noun} and the quantity-denoting noun the \textit{measure noun}.
Measure nouns can be all sorts of nouns which genuinely denote a quantity (such as \textit{litre} or \textit{amount}) but also nouns denoting containers, collections, etc. (such as \textit{glass} or \textit{bucket}).
In a similar vein, \citet[284]{Brems2003} considers nouns as measure nouns \textit{which, strictly speaking, do not designate a `measure', but display a more nebulous} (\textit{sic!}) \textit{potential for quantification} (see also \citealp[530]{Koptjevskaja2001}, and \citealp[338]{Rutkowski2007}).
For illustration purposes, in the English \textit{a cup of fine coffee}, \textit{cup} is the measure noun, and \textit{coffee} is the kind noun.

In the case at hand, three different syntactic configurations need to be distinguished \wrt case assignment inside German measure noun phrases.
If the kind noun forms an NP with a determiner, the construction resembles (and is usually called) a \textit{pseudo-partitive} (on partitives and pseudo-partitives see, e.\,g., \citealp{Barker1998,Selkirk1977,Stickney2007,Vos1999}; for a recent application of the terminology to German, see \citealp{Gerstenberger2015}).%
\footnote{If the kind noun is definite, the construction instantiates a true partitive.
Whereas partitives are constructions denoting a proper part-of relation as in \textit{a sip of the wine}, pseudo partitives -- albeit syntactically similar and diachronically related to partitives in many languages -- merely denote quantities and contain indefinite kind nouns as in \textit{a sip of wine}.
%In the literature on German, some authors incorrectly call the pseudo-partitive a \textit{partitive} \cite{Hentschel1993} while some realise the difference and at least mention it \citep{Eschenbach1994,GallmannLindauer1994,Loebel1989,Zimmer2015}.
}
Here, the kind noun is in the genitive, and I refer to the construction in (\ref{ex:intro:pseudopartitive1}) as the \textit{Pseudo-partitive Genitive Construction} (PGC).

@@

If the kind noun is bare -- \ie\ if it does neither come with a determiner nor a modifying adjective -- it has to agree in case with the measure noun, and the genitive seen in the PGC is not acceptable, see (\ref{ex:intro:narrowapposition}).

@@

This construction is usually classified as a \textit{Narrow Apposition Construction} \citep{Loebel1986}, henceforth NAC.%
\footnote{The construction as in (\ref{ex:intro:narrowapposition2}) is also referred to as the \textit{Direct Partitive Construction} for other Germanic languages in which the PGC with the synthetic genitive is not available.
This nomenclature makes sense in contrast to the \textit{Indirect Partitive Construction} with prepositional linkers translating to `of' -- \ie\ analytic genitives -- in such languages, see \cite{HankamerMikkelsen2008} for Danish.
For German, this terminology is not distinctive enough, which is why I use the terms NAC and PGC.}
Notice that the unavailability of the genitive on the kind noun can be seen as following from a rather quirky constraint that genitive NPs in German require the presence of some strongly case-marked element (determiner or adjective) in addition to the head noun in order to be acceptable \citep{GallmannLindauer1994,Schachtl1989}.
While only an accusative is shown in (\ref{ex:intro:narrowapposition}), the kind noun obligatory agrees in case also with nominative and dative measure nouns.

The actual alternation can be observed \textit{only when the kind noun occurs with an attributive adjective but without a determiner}, as in (\ref{ex:intro:alternation}), where both the NAC in (\ref{ex:intro:alternation1}) and the PGC in (\ref{ex:intro:alternation2}) are equally acceptable.
They are by-and-large functionally and semantically equivalent.
However, Section~\ref{sec:analyses} is devoted to developing hypotheses about subtle differences between them.%
\footnote{Some descriptive and normative grammars take stronger positions w.\,r.\,t.\ the acceptability of the two options.
See \cite{Hentschel1993,Zimmer2015} for analyses of the sometimes absurd stances taken in grammars of German.
As the usage and experimental data presented below (especially Section~\ref{sec:exp:fc}) should render it clear, there might be preferences under certain circumstances, but we cannot assume either construction to be unacceptable.
}

@@

\begin{table}
  \centering
  \begin{tabular}{lccc}
    \multicolumn{1}{r}{kind NP is:} & bare noun NP & NP with adjective & NP with determiner \\
    & [\ldots N\Subsf{meas} [N\Subsf{kind}]] & [\ldots N\Subsf{meas} [AP N\Subsf{kind}]] & [\ldots N\Subsf{meas} [D N\Subsf{kind}]] \\
    \midrule
    narrow apposition         & \NACb & \NACa & ---   \\
    pseudo-partitive genitive & ---   & \PGCa & \PGCd \\
  \end{tabular}
  \caption{Distribution of the NAC and PGC constructions in different NP structures}
  \label{tab:constructions}
\end{table}

The distribution of the case patterns in the NAC (case identity between the measure noun and the kind noun) and in the PGC (genitive on the kind noun, regardless of the case of the measure noun) depending on the structure of the kind NP are summarised in Table~\ref{tab:constructions}.
I call the narrow apposition construction with a bare kind noun the \NACb, the partitive genitive with a determiner in the kind noun phrase the \PGCd.
I call the alternating variants with an adjective but no determiner in the kind noun phrase \NACa, and \PGCa, respectively.

@@

\begin{enumerate}[i.]
  \item{\label{it:intro:idiot} alleged alternatives to the NAC and the PGC}
  \item{\label{it:intro:datsg} alternative \textit{weak} forms of dative singular neuter adjectives}
%  \item{\label{it:intro:femsg} case syncretism in feminine NPs}
  \item{\label{it:intro:plurl} similar constructions with plural/collective kind nouns}
  \item{\label{it:intro:noifl} grammaticalised non-inflected measure nouns}
  \item{\label{it:intro:other} alternative constructions for expressing quantities}
\end{enumerate}

@@

As for (\ref{it:intro:datsg}), a complication with neuter kind nouns in the dative is mentioned by \citet[20--22]{Zimmer2015}.
In the \NACa, the adjective normally inflects with the so-called \textit{strong} case and number suffixes, which are used when no determiner with strong inflection is present in the NP.
\cite{Zimmer2015} reports a high number of occurrences of adjectives being inflected with the \textit{weak} inflectional marker, which are normally only used if a strongly inflected determiner precedes the adjective: \textit{kalt-en Wasser} `cold water' instead of \textit{kalt-em Wasser}.%
\footnote{See Section~\ref{sec:analyses} for more discussion of adjectival inflection.}
In my corpus sample, this tendency was not nearly as clear, and there was a high number of very noisy sentences among those potentially showing this pattern.
Hence, I do not discuss these forms.%
\footnote{All corpus data and scripts used for this paper will be released freely, so further examination of the few cases maybe showing this kind of inflection is possible.}

% \label{page:femininesyncretism}
% With feminine kind nouns, (\ref{it:intro:femsg}) is relevant.
% Whereas the singular masculine and neuter nominal subsystem is a four case system (at least for NPs with adjectives and determiners), feminine singular NPs show nominative-accusative and dative-genitive syncretisms.
% Table~\ref{tab:syncretisms} shows this as well as the partial syncretism in the plural, where no gender distinctions are made.
% This means that with dative measure nouns, the alternation is unobservable if the kind noun is feminine, and it is problematic to speak of a \textit{genitive} kind noun construction in this case.
% This is taken into account in the statistical modelling and the discussion presented in Section~\ref{sec:corpusstudies}.
% 
% \begin{table}
%   \centering
%   \begin{tabular}{llll}
%      & Masculine\slash Neuter & Feminine & Plural \\
%      \midrule
%      nominative & rot-er Wein    & \multirow{2}{*}{frisch-e Sahne}   & \multirow{2}{*}{klein-e Äpfel} \\
%      accusative & rot-en Wein    &                                   &                                \\
%      dative     & rot-em Wein    & \multirow{2}{*}{frisch-er Sahne}  & klein-en Äpfel-n               \\
%      genitive   & rot-en Wein-es &                                   & klein-er Äpfel                 \\
%   \end{tabular}
%   \caption{Case syncretisms in NPs with an adjective and without a determiner (\textit{roter Wein} `red wine', \textit{frische Sahne} `fresh cream', \textit{kleine Äpfel} `small apples')}
%   \label{tab:syncretisms}
% \end{table}

Turning to (\ref{it:intro:plurl}), readers might have noticed that so far only MNPs denoting quantities of substances (mass kind nouns such as \textit{ein Glas roter Wein\slash roten Weines} `a glass of red wine') have been discussed.
If the kind noun is a plural count noun as in \textit{ein Sack kleine Äpfel} (\NACa) or \textit{ein Sack kleiner Äpfel} (\PGCa) `a bag of small apples', a similar alternation between PGC and NAC can be observed.
In line with experimental results reported in \citet[15--16]{Zimmer2015}, I found that the PGC is so dominant with plural kind nouns (67 of 861 cases or over 92\%, cf.\ Section~\ref{sec:corpusstudies}) that the alternation cannot be analysed in the same way as in the singular case.%
\footnote{More concretely, I these cases were included in the regression analysis of the corpus data along with a factor encoding the number of the kind noun, this factor would most assuredly override any other regressor for the data points with a plural kind noun.}
While this will play a role in the interpretation of the corpus findings, MNPs with plural kind nouns will not be included in the corpus study and the experiments reported here.

As for (\ref{it:intro:noifl}), some measure nouns have been grammaticalised in a way that they always appear non-inflected.
They are typical measure nouns like \textit{Gramm} `gram', \textit{Pfund} `pound' or \textit{Prozent} `percent', which do not have plural forms at all.
%\footnote{With neglectable frequency, this effect is extended to container nouns like \textit{Glas} `glass', but with a twist.
%A non-inflected variant \textit{zwei Glas roter Wein} `two glasses of wine' and an inflected variant \textit{zwei Gläser roter Wein} co-exist.
%The non-inflected variant denotes the quantity of wine fitting into two glasses, the inflected variant denotes two glasses filled with wine.}
I treat these cases like other measure nouns because they enter into both the \NACa\ as in (\ref{ex:uninflectedmeasures:a}) and the \PGCa\ as in (\ref{ex:uninflectedmeasures:b}).
In Section~\ref{sec:analyses}, degrees of grammaticalisation as a factor influencing the alternation will be discussed, however.

@@

Finally, (\ref{it:intro:other}) suggests that there are alternative ways of expressing similar quantificational meanings.
In the variationist tradition, which is strongly influenced by Labovian sociolinguistics, the \textit{principle of accountability} would dictate that proper studies should examine a variationist \textit{variable}, \ie\ all different \textit{ways of saying the same thing} (\citealp{Labov1966}, \citealp{Labov1969}, for an overview see \citealp{Tagliamonte2012}).
In the case at hand, the variable might be something like \textit{measuring substances and collections}.
I argue that it is fully justified to focus narrowly on \NACa\ and \PGCa\ with their well-defined morph-syntactic properties, mostly because the alternative constructions are not used in the same range of contexts.
Two major alternatives might be considered.
First of all, the analytic (pseudo-)partitive with \textit{von} `of' is only available as an alternative to the PGC if the kind noun phrase contains a (definite or indefinite) determiner as in (\ref{ex:analyticalpartitive}).

\begin{exe}
  \ex\label{ex:analyticalpartitive} 
  \begin{xlist}
    \ex[ ]{\label{ex:analyticalpartitive:a}\gll ein Glas von dem roten Wein\\
                                                a glass of the red wine\\
                                                \glt a glass of the red wine
       }
    \ex[*]{\label{ex:analyticalpartitive:b}\gll ein Glas von rotem Wein\\
                                                a glass of red wine\\
                                                \glt a glass of red wine
       }
  \end{xlist}
\end{exe}

This means that the \textit{von} (pseudo-)partitive does not compete with the \NACa\ and the \PGCa.
In fact, there is not a single context in which they can be interchanged.

Second constructions with \textit{voll} (\textit{von})\slash\textit{voller} `full (of)' and \textit{mit} `with' are available, see (\ref{ex:alternatives}).

@@

These construction have very idiosyncratic properties (the construction with \textit{voller} is discussed in \citealp{Zeldes2018}, the construction with \textit{mit} is discussed in \citealp{Bhatt1990}), they are not semantically equivalent to the NAC and the PGC, and they are only available for a small subset of measure nouns.
All of these constructions are only used with measure nouns denoting containers, and the whole NP always refers to the container, and never to the quantity contained in it.
They are consequently incompatible with measure nouns denoting natural portions such as \textit{Schluck} `gulp' or \textit{Haufen} `heap' and strongly grammaticalised nouns such as \textit{Gramm} `gram'.
This disqualifies them as alternatives to the \NACa\ or \PGCa\ in most contexts, and they are thus decidedly \textit{not} more ways of saying the same thing. 
It is therefore reasonable to focus on the two well-defined variants.

@@

This section briefly reviews existing analyses of the \PGCa vs.\ \NACa\ alternation and related issues.
I also develop my own analysis and the appropriate hypotheses for the empirical studies presented in Sections~\ref{sec:corpusstudies} and \ref{sec:externalvalidation}.
I discuss four main factors that might influence the alternation:
first, the different syntactic structures of the variants \NACa\ and \PGCa\ in relation to their syntactic prototypes \NACb\ and \PGCd;
second, degrees of grammaticalisation of the measure noun associated with the two prototypes;
third, a semantic prototypicality effect of the degrees of grammaticalisation (plurality);
fourth, preferences in different registers.

The first aspect is related to analyses of the morpho-syntax of German MNPs.
A major interest in descriptive and generative linguistics is to find a plausible analysis of the hierarchical syntactic structure of the constructions in question.
The first question is which noun constitutes the head of the whole construction.
This was answered quite clearly by \cite{Loebel1986} already (see also \citealp[213]{Eschenbach1994}, and \citealp[16]{GallmannLindauer1994}).
Subject-verb agreement is always realised on the measure noun, and we can therefore assume that the measure noun is the head.
However, the MNP-internal structure has more interesting cues to offer if the strong\slash weak inflection patterns of adjectives are taken into account.
In NPs with a strongly inflected determiner, attributive adjectives inflect according to the massively syncretistic \textit{weak} pattern.
%\footnote{The weak pattern has only two suffixes.
%The structural cases (nominative and accusative) in the singular are marked with \textit{-e}, the oblique cases (dative and genitive) as well as the entire plural are marked with \textit{-en}.
%The only exception to this rule is the masculine accusative singular, which is also marked with \textit{-en}.}
If there is no determiner (as is the case in the alternating constructions), attributive adjectives inflect like determiners themselves, however.
This is called the \textit{strong} inflectional pattern.
Thus, the adjectives in the \NACa\ and the \PGCa\ have properties of adjectives as well as determiners.
They are lexical adjectives and function as attributive modifiers.
On the other hand, they are inflected like determiners, and they are the leftmost element in the NP, which is typical of determiners.
This unusual double nature of adjectives in NPs without determiners leads to a plausible probabilistic interpretation of the pattern shown in Table~\ref{tab:constructions}.
Whenever speakers classify the adjective in the kind noun phrase more as a determiner, they have to use the \PGCa\ because, if there is a determiner, the PGC (just like in the case of the \PGCd) is the only option.
When they classify the adjective more as an adjective, the kind NP has no determiner, and they have to use the \NACa\ (like in the case of the \NACb).%
\footnote{While the generative analysis presented in \cite{Bhatt1990} cannot properly deal with probabilistic effects, Bhatt comes close to this interpretation by analysing the kind NP in the GPC as a DP and in the NAC as an NP.}
In a framework that assumes similarity-based probabilistic classification, such as Prototype Theory, we can regard the \PGCd\ as instantiating the PGC prototype and the \NACb\ as the NAC prototype.
The choice between \PGCa\ and \NACa\ is then expected to be influenced by the degree of similarity to either of these prototypes.
I will argue below that the prototypes differ mainly in that they attract different (classes of) measure nouns more or less strongly, mostly because of degrees of grammaticalisation.
However, if lemmas are attracted with different strengths by the two prototypes, then we can expect a baseline attraction effect:
nouns which occur more often in the \PGCd\ should favour the \PGCa, and nouns which occur more often in the \NACb\ should favour the \NACa.
Put differently, a noun's relative frequencies in the prototypical non-alternating constructions should be indicative of same noun's preference to appear in the two alternating variants.
This baseline attraction effect will be quantified in the corpus study Section~\ref{sec:corpusstudies}.
